Appendix: A fragile guardrail---The minimal effect of election fraud on protest intensity in non-democracies

### 2025-11-14 Output/authoritarian_only_pdf_pop_rev.Rmd
Shaggy additions to research design section on endogeneity and omitted variables.

---
output:
  pdf_document:
linkcolor: blue
bibliography: C:/Users/colej/Documents/Research projects/bib_ch.bib
#biblio-style: "apalike"
fontsize: 12pt
link-citations: yes
toc: false
csl: C:/Users/colej/Documents/Research projects/ji_manipulation_protest/american-journal-of-political-science.csl
indent: true
abstract: "Two models of the relationship between election manipulation and protest can be found in the literature. One approach holds that election fraud reveals an incumbent who is more vulnerable to protest; in this view, protest risk can result in cleaner elections. An alternative approach contends that election manipulation can be a sign of underlying regime strengths; under this framework, protest risk is less likely to deter manipulation. This article interrogates the assumptions underlying these models, and conducts a novel test of their implications. Using data on the duration of protest waves in 659 authoritarian cases from 1990-2012, this paper shows that governments which manipulate elections more also defeat protests more quickly. This supports the ‘regime-strength’ model, and helps explain why electoral authoritarianism is a sticky equilibrium amid ongoing democratic backsliding. It suggests researchers should look beyond protest risk in models of electoral manipulation."
header-includes: 
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[table]{labelformat=empty}
  - \usepackage{multirow}
  - \usepackage{lscape}
  - \usepackage{pdflscape}
  - \usepackage{longtable}

@@

@@

Specifically, the results highlight patterns that might be considered puzzling, whichever model of electoral manipulation we prefer. For the regime-weakness model, it is potentially puzzling that most manipulated elections go un-protested. Regime-weakness models may explain this pattern by arguing that, most of the time, incumbents produce the 'right' amount of manipulation to avoid protest in equilibrium. Yet the results of this study challenge that interpretation, since it demonstrates that outside this equilibrium (i.e. when elections are rigged and the opposition protests), higher levels of manipulation are associated with greater regime strength, not less.

In rejoinder, it might reasonably be wondered why protests occur after highly manipulated elections, if manipulation is a signal of strength? First, in my view, this interpretation unnecessarily exports the instrumental logic of protest from the regime-weakness model to the regime-strength model. As discussed earlier in the paper, instrumental calculations about overthrowing incumbents are but one component of a long and varied literature on protest, which has identified multiple other expressive and strategic motives for protest. These other motives may foster protest movements in the aftermath of rigged elections, even when severe electoral manipulation indicates considerable regime strength. Second, if manipulation does indeed indicate strength, incumbents have every incentive to build and deploy those strengths against stronger opposition movements (who will themselves be more capable of initiating protest).

>>>>

---
output:
  pdf_document:
linkcolor: blue
bibliography: C:/Users/colej/Documents/Research projects/bib_ch.bib
#biblio-style: "apalike"
fontsize: 12pt
link-citations: yes
toc: false
csl: C:/Users/colej/Documents/Research projects/ji_manipulation_protest/american-journal-of-political-science.csl
indent: true
abstract: "Two models of the relationship between election manipulation and protest can be found in the literature. One approach holds that more severe election fraud reveals an incumbent who is more vulnerable to protest; in this view, protest risk can result in cleaner elections. An alternative approach contends that election manipulation can be a sign of underlying regime strengths; under this framework, protest risk is less likely to deter manipulation. This article interrogates the assumptions underlying these models, and conducts a novel test of their implications. Using data on the occurrence and duration of protest waves in 659 authoritarian cases from 1990-2012, this paper shows that governments which manipulate elections more also defeat protests more quickly. This supports the ‘regime-strength’ model, and helps explain why electoral authoritarianism is a sticky equilibrium amid ongoing democratic backsliding. It suggests researchers should look beyond protest risk in models of electoral manipulation."
header-includes: 
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[table]{labelformat=empty}
  - \usepackage{multirow}
  - \usepackage{lscape}
  - \usepackage{pdflscape}
  - \usepackage{longtable}

@@

Modeling protest duration rather than initiation helps reduce concerns about endogeneity. For instance, some regime-weakness models expect that incumbents will calibrate the level of election fraud to minimize risk of protest initiation. In that case, strong incumbents would manipulate more and weak incumbents less, but neither would systematically generate protest. Unpredictable and likely statistically insignificant correlations between fraud and protest initiation would be expected. Focusing on duration partially avoids this risk. Once protests begin (whether instrumental or multi-purpose), the die is cast. Authoritarian incumbents have strong incentives to defeat a public challenge to their legitimacy, while committed opponents have similar incentives to stand firm. Both should be expected to put significant effort into defeating / maintaining the protest movements, as a general rule, and the process by which the challenge is settled helps reveal previously hidden information about each side. Whether the incumbent manipulated 'too much' or 'too little' is now beside the point, and it is possible to observe whether incumbents who produce extensive election fraud are stronger or weaker than those that do not.

While strategic endogeneity is less of a concern for this research design, selection bias and confounding variables may still be. For instance, it could be the case that incumbents facing well-organized opposition-oriented civil society may manipulate less; at the same time, such well-organized opposition may be able to sustain protest movements longer. A similar logic might imply that incumbents should manipulate less when economic grievances are high (for fear of sparking protest), and that economic troubles might make protests more durable. Either case could produce the relationship expected in H2, but could be considered spurious. To reduce this risk, I include several potential controls for opposition strength, and for grievance. Controlling for these variables helps isolate the relationship between manipulation severity and protest duration, holding opposition strength and grievances constant.

As such, the results should be considered causally identified only to the extent that unknown or unobservable variables are not confounding the relationship between fraud and protest duration. While this problem is difficult to surmount with observational data, this research design offers three defenses. First, I control for known possible confounders, including the occurrence of a successful electoral protest in a nearby country (which might frighten incumbents into manipulating less and inspire opponents to hold the line longer). Second, I present two analyses in the appendix, which show that increased risk of protest leads incumbents to *intensify* their manipulation efforts, rather than diminish them. Thus a negative relationship between fraud and protest duration should indicate the direct hypothesized effect, rather than one that is driven by relatively weaker incumbents avoiding manipulation. Third, and relatedly, the research design pits two incompatible hypotheses against one another, rather than a single hypothesis against the null. A null relationship between fraud and protest duration might not be strong evidence against the regime-weakness model, but a negative relationship should be (after accounting for necessary controls).

@@

Specifically, the results highlight patterns that might be considered puzzling, whichever model of electoral manipulation we prefer. For the regime-weakness model, it is potentially puzzling that most manipulated elections go un-protested. Regime-weakness models may explain this pattern by arguing that, most of the time, incumbents produce the 'right' amount of manipulation to avoid protest in equilibrium. Such careful calibration is difficult to ensure, however, when incumbents must rely on thousands of low-level agents to produce the 'right' amount of manipulation; these agents may have collective incentives to shirk and underproduce, or to demonstrate their loyalty by over-producing. [ADD CITES] The results of this study add an empirical challenge, since it demonstrates that when the deterrence equilibrium fails (i.e. when the incumbent rigs elections and the opposition protests), higher levels of manipulation are associated with greater regime strength, not less.

In rejoinder, it might reasonably be wondered why protests occur after highly manipulated elections, if manipulation is a signal of strength? First, in my view, this interpretation unnecessarily exports the instrumental logic of protest from the regime-weakness model to the regime-strength model. As discussed earlier in the paper, instrumental calculations about overthrowing incumbents are but one component of a long and varied literature on protest, which has identified multiple other expressive and strategic motives for protest. These other motives may foster protest movements in the aftermath of rigged elections, even when severe electoral manipulation indicates considerable regime strength. Second, if manipulation does indeed indicate strength, incumbents have every incentive to build and deploy those strengths in reaction to stronger opposition movements (who will themselves be more capable of initiating protest). Evidence that incumbents do, indeed, appear to increase manipulation in the face of increased protest threat is presented in the supplementary appendix.
